Transfer of Value Functions via Variational Methods

### 2018-05-18 NIPS 2018/main.tex
centered rooms image and legend

\begin{document}
\setlength{\abovedisplayskip}{0.5pt}
\setlength{\belowdisplayskip}{0.5pt} 
% \nipsfinalcopy is no longer used

@@

Let us observe that the distribution $\mathcal{D}$ over tasks induces a distribution over optimal $Q$-functions. \todo{Should we just make the assumption that Q-functions share knowledge in their weights?} Furthermore, for any MDP, learning its optimal $Q$-function is sufficient for solving the problem. Thus, one can safely replace the distribution over tasks with the distribution over their optimal value functions. In our parametric settings, we reduce the latter to a distribution $p(\bm{w})$ over weights.\\ 
Assume, for the moment, that we know the distribution $p(\bm{w})$ and consider a dataset $D=\left\{(s_i,a_i,s_i',r_i) \mid i = 1,2,\dots N\right\}$ of samples from some task $\mathcal{M}_\tau \sim \mathcal{D}$ that we want to solve. Then, we can compute the posterior distribution over weights given such dataset by applying Bayes theorem as $p(\bm{w} | D) \propto p(D | \bm{w})p(\bm{w})$. Unfortunately, this cannot be directly used in practice since we do not have a model of the likelihood $p(D|\bm{w})$. In such case, it is very common to make strong assumptions on the MDPs or the $Q$-functions to get tractable posteriors. However, in our transfer settings, all distributions involved depend on the family of tasks under consideration and making such assumptions is likely to limit the applicability to specific problems. Thus, we take a different approach to derive a more general, but still well-grounded, solution. Recall that our final goal is to move the total probability mass over the weights minimizing some empirical loss measure, which in our case is the TD error $\norm{B_{\bm{w}}}_D^2$. Then, given a prior $p(\bm{w})$, we know from PAC-Bayesian theory that the optimal Gibbs posterior $q$ takes the form (e.g., \cite{catoni2007pac}):
\begin{equation}\label{eq:gibbs}
q(\bm{w}) = \frac{e^{-\Lambda\norm{B_{\bm{w}}}_D^2}p(\bm{w})}{\int e^{-\Lambda\norm{B_{\bm{w}'}}_D^2}p(d\bm{w}')},
\end{equation}
for some parameter $\Lambda > 0$. \todo{Maybe say how this is obtained} Since $\Lambda$ is typically chosen to increase with the number of samples $N$, in the remaining, we set it to $\lambda^{-1}N$, for some constant $\lambda > 0$. Notice that, whenever the term $e^{-\Lambda\norm{B_{\bm{w}}}_D^2}$ can be interpreted as the actual likelihood of $D$, $q$ becomes a classic Bayesian posterior. Although we now have an appealing distribution, the integral at the denominator of \eqref{eq:gibbs} is intractable to compute even for simple $Q$-function models. Thus, we propose a variational approximation $q_{\bm{\xi}}$ by considering a simpler family of distributions parameterized by $\bm{\xi} \in \Xi$. Then, our problem reduces to finding the variational parameters $\bm{\xi}$ such that $q_{\bm{\xi}}$ minimizes the Kullback-Leibler (KL) divergence w.r.t. the Gibbs posterior $q$. From the theory of variational inference (e.g., \cite{blei2017variational}), this can be shown to be equivalent to minimizing the well-known (negative) \textit{evidence lower bound} (ELBO):
\begin{align} \label{eq:elbo}
\min_{\bm{\xi}\in\Xi} \mathcal{L}(\bm{\xi}) = \mathbb{E}_{\bm{w} \sim q_{\bm{\xi}}}\left[\norm{B_{\bm{w}}}_D^2\right] + \frac{\lambda}{N}KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w})\right).
\end{align}
The approximate posterior trades-off between placing probability mass over those weights $\bm{w}$ that have low expected TD error (first term), and staying close to the prior distribution (second term). Assuming that we can compute the gradients of \eqref{eq:elbo} w.r.t. the variational parameters $\bm{\xi}$, our objective can be optimized using any stochastic optimization algorithm, as shown in the next subsections.

\begin{algorithm}[t]
\begin{algorithmic}[1]
\REQUIRE Target task $\mathcal{M}_\tau$, source $Q$-function weights $\mathcal{W}_s$, batch size $M$
% \\\hrulefill

@@

\STATE{Sample initial state: $s_0 \sim p_0$}

\WHILE{$s_h$ is not terminal}
\STATE{Sample weights: $\bm{w} \sim q_{\bm{\xi}}(\bm{w})$}
\STATE{Take action $a_h = \argmax_a Q_{\bm{w}}(s_h,a)$}
\STATE{Observe transition $s_{h+1} \sim \mathcal{P}_{\tau}(\cdot | s_h,a_h)$ and collect reward $r_{h+1} = \mathcal{R}_{\tau}(s_h,a_h)$}
\STATE{Add sample to the replay buffer: $D \leftarrow D \cup \langle s_h,a_h,r_{h+1},s_{h+1} \rangle$}
\STATE{Sample mini-batch $D' = \langle s_i,a_i,r_i,s_i' \rangle_{i=1}^{M}$ from $D$}
\STATE{Estimate the gradient $\nabla_{\bm{\xi}}\mathcal{L}(\bm{\xi})$ using $D'$}
\STATE{Update $\bm{\xi}$ in the direction of $-\nabla_{\bm{\xi}}\mathcal{L}(\bm{\xi})$ using any stochastic optimizer (e.g., ADAM)}
\ENDWHILE
\UNTIL{forever}
\end{algorithmic}
\caption{Variational Transfer} \label{alg:transfer}
\end{algorithm}
We now highlight our general transfer procedure in Algorithm~\ref{alg:transfer}, while deferring a description of specific choices for the involved distributions to the next two subsections. Given a set of weights $\mathcal{W}_s$ from the source tasks' optimal $Q$-functions, we start by estimating the prior distribution (line 1), and we initialize the variational parameters by minimizing the KL divergence w.r.t. such distribution (line 2).\footnote{If the prior and approximate posterior were in the same family of distributions we could simply set $\bm{\xi}$ to the prior parameters. However, we are not making this assumption at this point.} Then, at each time step of interaction, we re-sample the weights from the current approximate posterior and act greedily w.r.t. the corresponding $Q$-function (lines 7,8). After collecting the new experience (lines 9-10), we draw a mini-batch of samples from the replay buffer (line 11), use this to estimate the objective function gradient (line 12), and update the variational parameters (line 13).\\
The key property of our approach is the weight resampling at line 7, which resembles the well-known Thompson sampling approach adopted in multi-armed bandits \cite{bubeck2012regret} and closely relates to the recent value function randomization \cite{osband2014generalization}. At each time we guess what the task we are trying to solve based on our current belief is and we act as if such guess were true. This mechanism allows an efficient adaptive exploration of the target task. Intuitively, during the first steps of interaction, the agent is very uncertain about the current task, and such uncertainty induces stochasticity in the chosen actions, allowing rather informed exploration to take place. Consider, for instance, that actions that are bad on average for all tasks are improbable to be sampled, while this cannot happen in uninformed exploration strategies, like $\epsilon$-greedy, before learning takes place. As the learning process goes on, the algorithm will quickly figure out which task is being faced, thus moving all the probability mass over the weights minimizing the TD error. From that point, sampling from the posterior is approximately equivalent to deterministically taking such weights, and no more exploration will be performed.\\
Finally, notice the generality of the proposed approach: as far as the objective $\mathcal{L}$ is differentiable in the variational parameters $\bm{\xi}$, and its gradients can be efficiently computed, any approximator for the $Q$-function and any prior/posterior distributions can be adopted. For the latter, we describe two practical choices in the next two sections.

@@

We now restrict to a specific choice of the prior and posterior families that makes our algorithm very efficient and easy to implement. We assume that optimal $Q$-functions (or better, their weights) follow a multivariate Gaussian distribution. That is, we model the prior as $p(\bm{w}) = \mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and we learn its parameters from the set of source weights using maximum likelihood estimation (with small regularization to make sure the covariance is positive definite). Then, our variational family is the set of all well-defined Gaussian distributions, i.e., the variational parameters are $\Xi = \left\{ (\bm{\mu},\bm{\Sigma}) \mid \bm{\mu}\in\mathbb{R}^d, \bm{\Sigma}\in\mathbb{R}^{d\times d}, \bm{\Sigma}\succ 0 \right\}$. To prevent the covariance from becoming not positive definite, we consider its Cholesky decomposition $\bm{\Sigma} = \bm{LL}^T$ and learn the lower-triangular Cholesky factor $\bm{L}$ instead. In this case, deriving the gradient of the objective is very simple. Both the KL between two multivariate Gaussians and its gradients have a simple closed-form expression. The expected log-likelihood, on the other hand, can be easily differentiated by adopting the reparameterization trick (e.g., \cite{hoffman2013stochastic,rezende2014stochastic}). We report these results in App. \ref{}.

@@

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This model resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distributions. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians.  Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances. Finally, since the KL divergence between two mixtures of Gaussians has no closed-form expression, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and the exact posterior. Among the many upper bounds available, we adopt the one proposed in \cite{hershey2007approximating} (see App. \ref{}).

@@

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B_{\bm{w}}}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This issue rarely is a problem since typical value-based algorithms are semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are quite different from common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum and aims only to adapt the weights in some direction of lower error so as to quickly converge to the solution of the target task. Unfortunately, this property does not hold for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This property is clearly undesirable in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. To differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}_{\bm{w}} = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}_{\bm{w}}}_D^2$, is now differentiable w.r.t. $\bm{w}$. Further considerations on how to better optimize it are given in App. \ref{}.

@@

\paragraph{The Rooms Problem} \label{sec:gw}

@@

\begin{figure*}[t] 
  \begin{subfigure}[b]{0.32\textwidth}
    %\includegraphics[trim=0.0cm 0.2cm 0.0cm 0.2cm,clip=true,height=2.6cm, width=2.0cm]{images/rooms2.pdf}
    \begin{tikzpicture}[scale=0.37]
\draw [white] (0,0) grid (11,8);
%\begin{scope}[shift={(0,2)}]
%\draw [thick] (4,5) arc [radius=1, start angle=180, end angle= 270];
%\path [fill=gray!50, very thin] (4,5) arc [radius=1, start angle=180, end angle= 270] -- (5,4) --(5,5)--(4,5);
%\draw [thick] (0,1) arc [radius=1, start angle=90, end angle= 0];
%\path [fill=lightgray!10, very thin] (0,1) arc [radius=1, start angle=90, end angle=0] -- (1,0) --(0,0)--(0,1);
%
%%\draw [fill=black!50] (1,1) circle [radius=0.1cm];
%%\node at (1,1) [ right] {$A$};
%
%\node at (1.8,-0.5)[left] {\color{darkgray}\small Start};
%\node at (3.1,5.5) [right]{\color{darkgray}\small Goal};
%
%\draw [thick, gray] (0,0) rectangle (5,5);
%\draw [line width=0.5mm, gray] (1.25,5/3) -- (5,5/3);
%\draw [line width=0.5mm, gray] (0,5/3) -- (0.75,5/3);
%\draw [line width=0.5mm, gray] (0,10/3) -- (3.25,10/3);
%\draw [line width=0.5mm, gray] (3.75,10/3) -- (5,10/3);

@@

\begin{scope}[shift={(0,3)}]
\begin{scope}[shift={(3,0)}]
\draw [thick] (4,5) arc [radius=1, start angle=180, end angle= 270];
\path [fill=gray!50, very thin] (4,5) arc [radius=1, start angle=180, end angle= 270] -- (5,4) --(5,5)--(4,5);
\draw [thick] (0,1) arc [radius=1, start angle=90, end angle= 0];
\path [fill=lightgray!10, very thin] (0,1) arc [radius=1, start angle=90, end angle=0] -- (1,0) --(0,0)--(0,1);

@@

\end{scope}

\draw [thick, line width=0.6mm, dotted, nt] (0,-1.5) -- (1,-1.5);
\node at (1.5,-1.5) [right] {NT};
\draw [thick, line width=0.6mm, dashed,gvt] (6, -1.5) -- (7,-1.5);
\node at (7,-1.5) [right] {GVT};
\draw [thick, line width=0.6mm, mgvt1] (0,-2.5) -- (1,-2.5);
\node at (1.5,-2.5) [right] {1-MGVT};
\draw [thick, line width=0.6mm, mgvt3] (6, -2.5) -- (7,-2.5);
\node at (11.2,-2.5) [left] {3-MGVT};

\end{scope}
\end{tikzpicture} 
    \caption{The rooms problem}\label{f:rooms}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0cm 0cm 1.2cm 1.1cm,clip=true,height=2.9cm]{images/three-room/lrev.pdf}
    \caption{Rooms: two doors moving}\label{f:two_doors}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0cm 0cm 1.2cm 1.1cm,clip=true,height=2.9cm]{images/three-room-gen/lrev.pdf}
    \caption{Rooms: one door moving}\label{f:one_door}
  \end{subfigure}
%    \begin{subfigure}[b]{0.27\textwidth}
%    \includegraphics[trim=0cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/likelihood/lrew.pdf}
%    \caption{}\label{f:likelihood}
%  \end{subfigure}

>>>>

\begin{document}
\setlength{\abovedisplayskip}{1.5pt}
\setlength{\belowdisplayskip}{1.5pt} 
% \nipsfinalcopy is no longer used

@@

Let us observe that the distribution $\mathcal{D}$ over tasks induces a distribution over optimal $Q$-functions. \todo{Should we just make the assumption that Q-functions share knowledge in their weights?} Furthermore, for any MDP, learning its optimal $Q$-function is sufficient for solving the problem. Thus, one can safely replace the distribution over tasks with the distribution over their optimal value functions. In our parametric settings, we reduce the latter to a distribution $p(\bm{w})$ over weights.

Assume, for the moment, that we know the distribution $p(\bm{w})$ and consider a dataset $D=\left\{(s_i,a_i,s_i',r_i) \mid i = 1,2,\dots N\right\}$ of samples from some task $\mathcal{M}_\tau \sim \mathcal{D}$ that we want to solve. Then, we can compute the posterior distribution over weights given such dataset by applying Bayes theorem as $p(\bm{w} | D) \propto p(D | \bm{w})p(\bm{w})$. Unfortunately, this cannot be directly used in practice since we do not have a model of the likelihood $p(D|\bm{w})$. In such case, it is very common to make strong assumptions on the MDPs or the $Q$-functions to get tractable posteriors. However, in our transfer settings, all distributions involved depend on the family of tasks under consideration and making such assumptions is likely to limit the applicability to specific problems. Thus, we take a different approach to derive a more general, but still well-grounded, solution. Recall that our final goal is to move the total probability mass over the weights minimizing some empirical loss measure, which in our case is the TD error $\norm{B_{\bm{w}}}_D^2$. Then, given a prior $p(\bm{w})$, we know from PAC-Bayesian theory that the optimal Gibbs posterior $q$ takes the form (e.g., \cite{catoni2007pac}):
\begin{equation}\label{eq:gibbs}
q(\bm{w}) = \frac{e^{-\Lambda\norm{B_{\bm{w}}}_D^2}p(\bm{w})}{\int e^{-\Lambda\norm{B_{\bm{w}'}}_D^2}p(d\bm{w}')},
\end{equation}
for some parameter $\Lambda > 0$. \todo{Maybe say how this is obtained} Since $\Lambda$ is typically chosen to increase with the number of samples $N$, in the remaining, we set it to $\lambda^{-1}N$, for some constant $\lambda > 0$. Notice that, whenever the term $e^{-\Lambda\norm{B_{\bm{w}}}_D^2}$ can be interpreted as the actual likelihood of $D$, $q$ becomes a classic Bayesian posterior. Although we now have an appealing distribution, the integral at the denominator of \eqref{eq:gibbs} is intractable to compute even for simple $Q$-function models. Thus, we propose a variational approximation $q_{\bm{\xi}}$ by considering a simpler family of distributions parameterized by $\bm{\xi} \in \Xi$. Then, our problem reduces to finding the variational parameters $\bm{\xi}$ such that $q_{\bm{\xi}}$ minimizes the Kullback-Leibler (KL) divergence w.r.t. the Gibbs posterior $q$. From the theory of variational inference (e.g., \cite{blei2017variational}), this can be shown to be equivalent to minimizing the well-known (negative) \textit{evidence lower bound} (ELBO):
\begin{align} \label{eq:elbo}
\min_{\bm{\xi}\in\Xi} \mathcal{L}(\bm{\xi}) = \mathbb{E}_{\bm{w} \sim q_{\bm{\xi}}}\left[\norm{B_{\bm{w}}}_D^2\right] + \frac{\lambda}{N}KL\left(q_{\bm{\xi}}(\bm{w})\ ||\ p(\bm{w})\right).
\end{align}
The approximate posterior trades-off between placing probability mass over those weights $\bm{w}$ that have low expected TD error (first term), and staying close to the prior distribution (second term). Assuming that we can compute the gradients of \eqref{eq:elbo} w.r.t. the variational parameters $\bm{\xi}$, our objective can be optimized using any stochastic optimization algorithm, as shown in the next subsections.

\begin{algorithm}[t]
\begin{algorithmic}[1]
\REQUIRE Target task $\tau$, source $Q$-function weights $\mathcal{W}_s$, batch size $M$
\\\hrulefill

@@

\STATE{Sample initial state: $s_0 \sim p_0^{(\tau)}$}

\WHILE{$s_h$ is not terminal}
\STATE{Sample weights: $\bm{w} \sim q_{\bm{\xi}}(\bm{w})$}
\STATE{Take action $a_h = \argmax_a Q_{\bm{w}}(s_h,a)$}
\STATE{Observe transition $s_{h+1} \sim \mathcal{P}^{(\tau)}(\cdot | s_h,a_h)$ and collect reward $r_h = \mathcal{R}^{(\tau)}(s_h,a_h)$}
\STATE{Add sample to the replay buffer: $D \leftarrow D \cup \langle s_h,a_h,r_h,s_{h+1} \rangle$}
\STATE{Sample mini-batch $D' = \langle s_i,a_i,r_i,s_i' \rangle_{i=1}^{M}$ from $D$}
\STATE{Estimate the gradient $\nabla_{\bm{\xi}}\mathcal{L}(\bm{\xi})$ using $D'$}
\STATE{Update $\bm{\xi}$ in the direction of $-\nabla_{\bm{\xi}}\mathcal{L}(\bm{\xi})$ using any stochastic optimizer (e.g., ADAM)}
\ENDWHILE

\UNTIL{forever}
\end{algorithmic}
\caption{Variational Transfer} \label{alg:transfer}
\end{algorithm}

We now highlight our general transfer procedure in Algorithm~\ref{alg:transfer}, while deferring a description of specific choices for the involved distributions to the next two subsections. Given a set of weights $\mathcal{W}_s$ from the source tasks' optimal $Q$-functions, we start by estimating the prior distribution (line 1), and we initialize the variational parameters by minimizing the KL divergence w.r.t. such distribution (line 2).\footnote{If the prior and approximate posterior were in the same family of distributions we could simply set $\bm{\xi}$ to the prior parameters. However, we are not making this assumption at this point.} Then, at each time step of interaction, we re-sample the weights from the current approximate posterior and act greedily w.r.t. the corresponding $Q$-function (lines 7,8). After collecting and storing the new experience (lines 9-10), we draw a mini-batch of samples from the replay buffer (line 11), use this to estimate the objective function gradient (line 12), and, finally, update the variational parameters (line 13).

The key property of our approach is the weight resampling at line 7, which resembles the well-known Thompson sampling approach adopted in multi-armed bandits \cite{bubeck2012regret} and closely relates to the recent value function randomization \cite{osband2014generalization}. At each time we guess what the task we are trying to solve based on our current belief is and we act as if such guess were true. This mechanism allows an efficient adaptive exploration of the target task. Intuitively, during the first steps of interaction, the agent is very uncertain about the current task, and such uncertainty induces stochasticity in the chosen actions, allowing rather informed exploration to take place. Consider, for instance, that actions that are bad on average for all tasks are improbable to be sampled, while this cannot happen in uninformed exploration strategies, like $\epsilon$-greedy, before learning takes place. As the learning process goes on, the algorithm will quickly figure out which task is being faced, thus moving all the probability mass over the weights minimizing the TD error. From that point, sampling from the posterior is approximately equivalent to deterministically taking such weights, and no more exploration will be performed.

Finally, notice the generality of the proposed approach: as far as the objective $\mathcal{L}$ is differentiable in the variational parameters $\bm{\xi}$, and its gradients can be efficiently computed, any approximator for the $Q$-function and any prior/posterior distributions can be adopted. For the latter, we describe two practical choices in the next two sections.

@@

We now restrict to a specific choice of the prior and posterior families that makes our algorithm very efficient and easy to implement. We assume that optimal $Q$-functions (or better, their weights) follow a multivariate Gaussian distribution. That is, we model the prior as $p(\bm{w}) = \mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and we learn its parameters from the set of source weights using maximum likelihood estimation (with small regularization to make sure the covariance is positive definite). Then, our variational family is the set of all well-defined Gaussian distributions, i.e., the variational parameters are $\Xi = \left\{ (\bm{\mu},\bm{\Sigma}) \mid \bm{\mu}\in\mathbb{R}^d, \bm{\Sigma}\in\mathbb{R}^{d\times d}, \bm{\Sigma}\succ 0 \right\}$. To prevent the covariance from becoming not positive definite, we consider its Cholesky decomposition $\bm{\Sigma} = \bm{LL}^T$ and learn the lower-triangular Cholesky factor $\bm{L}$ instead. In this case, deriving the gradient of the objective is very simple. Both the KL between two multivariate Gaussians and its gradients have a simple closed-form expression. The expected log-likelihood, on the other hand, can be easily differentiated by adopting the reparameterization trick (e.g., \cite{hoffman2013stochastic,rezende2014stochastic}). Although these results are well-known in the literature, we report them in App. \ref{} to have a more self-contained description.

@@

Given the source tasks' weights $\mathcal{W}_s$, we model our estimated prior as a mixture of Gaussians with one equally weighted isotropic Gaussian centered at each weight: $p(\bm{w}) = \frac{1}{\abs{\mathcal{W}_s}}\sum_{\bm{w}_s \in \mathcal{W}_s}\mathcal{N}(\bm{w}|\bm{w}_s,\sigma_p^2\bm{I})$. This model resembles a kernel density estimator \cite{scott2015multivariate} and, due to its nonparametric nature, it allows capturing arbitrarily complex distributions. Consistently with the prior, we model our approximate posterior as a mixture of Gaussians.  Using $C$ components, our posterior is $q_{\bm{\xi}}(\bm{w}) = \frac{1}{C}\sum_{i=1}^C\mathcal{N}(\bm{w}|\bm{\mu}_i,\bm{\Sigma}_i)$, with variational parameters $\bm{\xi} = (\bm{\mu}_1,\dots,\bm{\mu}_C,\bm{\Sigma}_1,\dots,\bm{\Sigma}_C)$. Once again, we learn Cholesky factors instead of full covariances. Finally, since the KL divergence between two mixtures of Gaussians has no closed-form expression, we rely on an upper bound to such quantity, so that negative ELBO still upper bounds the KL between the approximate and the exact posterior. Among the many upper bounds available in the literature, we adopt the one proposed in \cite{hershey2007approximating}, which we report in App. \ref{} for the sake of completeness.

@@

From Sections \ref{sec:gvt} and \ref{sec:mgvt}, we know that differentiating the negative ELBO $\mathcal{L}$ w.r.t. $\bm{\xi}$ requires differentiating $\norm{B_{\bm{w}}}_D^2$ w.r.t. $\bm{w}$. Unfortunately, the TD error is well-known to be non-differentiable due to the presence of the max operator. This issue rarely is a problem since typical value-based algorithms are semi-gradient methods, i.e., they do not differentiate the targets (see, e.g., Chapter 11 of \cite{sutton1998reinforcement}). However, our transfer settings are quite different from common RL. In fact, our algorithm is likely to start from $Q$-functions that are very close to an optimum and aims only to adapt the weights in some direction of lower error so as to quickly converge to the solution of the target task. Unfortunately, this property does not hold for most semi-gradient algorithms. Even worse, many online RL algorithms combined with complex function approximators (e.g., DQNs) are well-known to be unstable, especially when approaching an optimum, and require many tricks and tuning to work well \cite{schaul2015prioritized,van2016deep}. This property is clearly undesirable in our case, as we only aim at adapting already good solutions. Thus, we consider using a residual gradient algorithm \cite{baird1995residual}. To differentiate the targets, we replace the optimal Bellman operator with the mellow Bellman operator introduced in \cite{asadi2017alternative}, which adopts a softened version of max called \textit{mellowmax}:
\begin{equation}
\mm_a Q_{\bm{w}}(s,a) = \frac{1}{\kappa} \log \frac{1}{|\mathcal{A}|}\sum_a e^{\kappa Q_{\bm{w}}(s,a)}
\end{equation}
where $\kappa$ is a hyperparameter and $|\mathcal{A}|$ is the number of actions. The mellow Bellman operator, which we denote as $\wt{T}$, has several appealing properties that make it suitable for our settings: (i) it converges to the maximum as $\kappa \rightarrow \infty$, (ii) it has a unique fixed point, and (iii) it is \textit{differentiable}. Denoting by $\wt{B}_{\bm{w}} = \wt{T}Q_{\bm{w}} - Q_{\bm{w}}$ the Bellman residual w.r.t. the mellow Bellman operator $\wt{T}$, we have that the corresponding TD error, $\norm{\wt{B}_{\bm{w}}}_D^2$, is now differentiable w.r.t. $\bm{w}$. Further considerations on how to better optimize it are given in App. \ref{}.

@@

\subsection{The Rooms Problem} \label{sec:gw}

We consider an agent navigating in the environment depicted in Figure~\ref{f:rooms}. The agent starts in the bottom-left corner and must move from one room to another to reach the goal position in the top-right corner. The rooms are connected by small doors whose locations are unknown to the agent. The state-space is modeled as a $10 \times 10$ continuous grid, while the action-space is the set of $4$ movement directions (up, right, down, left). After each action, the agent moves by $1$ in the chosen direction and the final position is corrupted by Gaussian noise $\mathcal{N}(0,0.2)$. In case the agent hits a wall, its position remains unchanged. The reward is $1$ when reaching the goal (after which the process terminates) and $0$ otherwise, while the discount factor is $\gamma = 0.99$. In this experiment, we consider linearly parameterized $Q$-functions with $121$ equally-spaced radial basis features.

@@

\begin{figure*}[t] 
  \begin{subfigure}[b]{0.32\textwidth}
    %\includegraphics[trim=0.0cm 0.2cm 0.0cm 0.2cm,clip=true,height=2.6cm, width=2.0cm]{images/rooms2.pdf}
    \begin{tikzpicture}[scale=0.37]
\draw [white] (0,0) grid (12,9);
%\begin{scope}[shift={(0,2)}]
%\draw [thick] (4,5) arc [radius=1, start angle=180, end angle= 270];
%\path [fill=gray!50, very thin] (4,5) arc [radius=1, start angle=180, end angle= 270] -- (5,4) --(5,5)--(4,5);
%\draw [thick] (0,1) arc [radius=1, start angle=90, end angle= 0];
%\path [fill=lightgray!10, very thin] (0,1) arc [radius=1, start angle=90, end angle=0] -- (1,0) --(0,0)--(0,1);
%
%%\draw [fill=black!50] (1,1) circle [radius=0.1cm];
%%\node at (1,1) [ right] {$A$};
%
%\node at (1.8,-0.5)[left] {\color{darkgray}\small Start};
%\node at (3.1,5.5) [right]{\color{darkgray}\small Goal};
%
%\draw [thick, gray] (0,0) rectangle (5,5);
%\draw [line width=0.5mm, gray] (1.25,5/3) -- (5,5/3);
%\draw [line width=0.5mm, gray] (0,5/3) -- (0.75,5/3);
%\draw [line width=0.5mm, gray] (0,10/3) -- (3.25,10/3);
%\draw [line width=0.5mm, gray] (3.75,10/3) -- (5,10/3);

@@

\begin{scope}[shift={(0,3.8)}]
\begin{scope}[shift={(4,0)}]
\draw [thick] (4,5) arc [radius=1, start angle=180, end angle= 270];
\path [fill=gray!50, very thin] (4,5) arc [radius=1, start angle=180, end angle= 270] -- (5,4) --(5,5)--(4,5);
\draw [thick] (0,1) arc [radius=1, start angle=90, end angle= 0];
\path [fill=lightgray!10, very thin] (0,1) arc [radius=1, start angle=90, end angle=0] -- (1,0) --(0,0)--(0,1);

@@

\end{scope}
\begin{scope}[shift={(1,0)}]
\draw [thick, line width=0.6mm, dotted, nt] (0,-1.5) -- (1,-1.5);
\node at (1.5,-1.5) [right] {\small NT};
\draw [thick, line width=0.6mm, dashed,gvt] (6, -1.5) -- (7,-1.5);
\node at (7,-1.5) [right] {\small GVT};
\draw [thick, line width=0.6mm, mgvt1] (0,-2.5) -- (1,-2.5);
\node at (1.5,-2.5) [right] {\small 1-MGVT};
\draw [thick, line width=0.6mm, mgvt3] (6, -2.5) -- (7,-2.5);
\node at (10.9,-2.5) [left] {\small 3-MGVT};
\end{scope}
\end{scope}
\end{tikzpicture} 
    %\caption{The rooms problem}\label{f:rooms}
  \end{subfigure}
  \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0cm 0cm 1.2cm 1.1cm,clip=true,height=2.9cm]{images/three-room/lrev.pdf}
    \caption{Rooms: two doors moving}\label{f:two_doors}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0cm 0cm 1.2cm 1.1cm,clip=true,height=2.9cm]{images/three-room-gen/lrev.pdf}
    \caption{Rooms: one door moving}\label{f:one_door}
  \end{subfigure}
%    \begin{subfigure}[b]{0.27\textwidth}
%    \includegraphics[trim=0cm 0cm 1.6cm 1.3cm,clip=true,height=2.9cm]{images/likelihood/lrew.pdf}
%    \caption{}\label{f:likelihood}
%  \end{subfigure}

	\begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/cartpole/lrev.pdf}
    \caption{Cartpole}\label{f:cartpole}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/mountaincar/lrev.pdf}
    \caption{Mountain Car}\label{f:mountaincar}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/maze/lrev_main.pdf}
    \caption{Maze Navigation}\label{f:maze_results}
  \end{subfigure}
  \caption{Expected Return as a function of the number of iterations averaged over $20$ independent runs. $95\%$ confidence intervals are shown.}
  \label{fig:ccmazeplots}

  \label{fig:dipfdbsgraphplots}
\end{figure*}

We generate a set of $50$ source tasks for the three-room environment of Figure~\ref{f:rooms} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in App. \ref{}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.

To better understand the differences between GVT and MGVT, we now consider transferring from a slightly different distribution than the one from which target tasks are drawn. We generate $50$ source tasks again but this time with the bottom door fixed at the center and the other one moving. Then, we repeat the previous experiment, allowing both doors to move when sampling target tasks. The results are shown in Figure~\ref{f:one_door}. Interestingly, MGVT seems almost unaffected by this change, proving that it has sufficient representation power to generalize to slightly different task distributions. The same does not hold for GVT, which now is not able to solve many of the sampled target tasks, as can be noticed from the higher variance. This result proves again that assuming Gaussian distributions can pose severe limitations in our transfer settings.
